pirouette: Determining the error BEAST2 makes in inferring a phylogeny

### 2019-03-05 article.tex
Add symbols to workflow figure

\giovanni{describe winner}

@@

\giovanni{describe winner}
\richel{indeed, after the figures also show it :-) }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h]
\centering
\begin{tabular}{ | r | l | l | l | l | }
\hline
\multirow{2}{*}{\textbf{Example}} & \multicolumn{2}{c|}{\textbf{True tree}} 
                                  & \multicolumn{2}{c|}{\textbf{Twin tree}} \\
\cline{2-5}
                                  & \textbf{median} & \textbf{range} & \textbf{median} & \textbf{range} \\
\hline
1  & $0.02$  & $0.0 - 0.27$  &         &               \\
2  & $0.02$  & $0.0 - 0.08$  &         &               \\
3  & $0.018$ & $0.0 - 0.1$   & $0.019$ & $0.0 - 0.11$  \\
4C & $0.22$  & $0.0 - 0.088$ & $0.009$ & $0.0 - 0.056$ \\
4G & $0.020$ & $0.0 - 0.075$ & $0.013$ & $0.0 - 0.052$ \\
\hline
\end{tabular}
\caption{Results of the examples. 
  For the true and twin tree,
  the median and range (from minimum to maximum value) of the errors 
  is shown. Experiment 4 is split in two rows
  by either the candidate ('C') or generative model ('G')
}
\label{tab:results}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\richel{check numbers}
\giovanni{
  I think that the current version of the discussion is too didascalic. 
  Most of what is said can be effectively summarized by table 5. 
  We can even add ESSes in it. 
}
\giovanni{
  Another point is that we should understand what is 
  the message we want to deliver in this discussion. 
  Since we are only using one tree we are not in the position to claim anything, 
  because a single tree is not an effective sample. 
  I saw that in some occasions you refer to the RQ as examples. 
  This might be the right way to proceed, 
  as our purpose is just to illustrate how to run all the possible experiments 
  on a full tree distribution. 
  I think we should make it clearer throughout the entire manuscript. 
  I am also interested in hearing Rampal's opinion on this.
}

\iffalse
Here we discuss how to interpret the results of \verb;pirouette;.
Figure \ref{fig:example_1} shows that the generative model 
gives an error distribution with a median of \richel{?}
The range of errors goes from \richel{?} to \richel{?}.
As we know the generative model, this error distribution is the minimum noise that BEAST2 produces for that generative model.
We can measure the effective samples as well (not shown), 
and all the Estimated Sample Sizes (ESSes) are above
the recommended value of 200 [\cite{drummond2015bayesian}], 
as shown in table \ref{tab:esses_example_1}.

Figure \ref{fig:example_2} shows that the best candidate model gives an error distribution with a median of \richel{?}. 
The range of errors goes from \richel{?} to \richel{?}. 
The inference model with the highest evidence follows a JC site model, RLN clock model and BD tree prior, with a weight of $0.675$. 
This is (unexpectedly) \giovanni{We are not using distributions, so we are totally subject to stochasticity. Nothing can be expected from just one tree.} better than the generative
model with a weight of $0.166$. 
The complete list of all evidences is shown in table \ref{tab:evidences_example_2}.
This means that it is rational to use this inference model to 
infer a posterior phylogeny distribution, as is done by \verb;pirouette;.
Also here, the ESSes are well above 200, 
as shown in table \ref{tab:esses_example_2}.

Figure \ref{fig:example_3} shows that the hand-picked model 
gives an error distribution with a median of \richel{?}, 
The twin tree has an error distribution with a median of \richel{?},
with a range from 0.0 to \richel{?}, which is an indication of the
error made by BEAST on that tree prior.
As there is little difference between the true and twin tree,
the (unknown) diversification model has little impact on the error
BEAST2 makes, hinting that the constant-rate birth-death model is
sufficiently good enough to explain the true phylogeny. 
The diversification model apparently has minor impact and need not to be
added to BEAST2.
We can draw this conclusion with confidence, as the ESSes are well above 200, 
as shown in table \ref{tab:esses_example_3}.

Figure \ref{fig:example_4} shows that \richel{?}
and 
gives an error distribution with a median of \richel{?}, 
The twin tree has an error distribution with a median of \richel{?},
with a range from 0.0 to \richel{?}, which is an indication of the
error made by BEAST on that tree prior.
The inference model with the highest evidence follows a JC site model, 
RLN clock model and BD tree prior, with a weight of ?. 
The complete list of all evidences is shown in table \ref{tab:evidences_example_4}.
As there is little difference between the true and twin tree,
the (unknown) diversification model has little impact on the error
BEAST2 makes, hinting that the constant-rate birth-death model is
sufficiently good enough to explain the true phylogeny. 
The diversification model apparently has minor impact and need not to be
added to BEAST2.
We can draw this conclusion with confidence \giovanni{no, we cannot: one tree is not enough}, as the ESSes are well above 200, 
as shown in table \ref{tab:esses_example_4}.
\fi

\giovanni{
  This is a first draft on how I think the discussion should be setup. 
  In this context there is no much need for the results in the table. 
  Let me know what you think about it.
}

In the four example research questions we showed how to use \verb;pirouette; to assess if BEAST2 is able, with a birth-death prior, to recover the original tree simulated by a Yule process. In principle any other, possibly much more complex, generative prior can be tested.
All tests were performed on a single original tree, hence not providing us enough statistical power to evaluate the level of accuracy of the inference. However if the same procedure were repeated and performed on a distribution with a sufficient number of generative trees, it could actually constitute a quantitative assessment of the quality of the inference, as shown in figures \ref{fig:example_1}, \ref{fig:example_2}, \ref{fig:example_3}, \ref{fig:example_4}. Importantly, \verb;pirouette; can directly show the comparison between the results obtained on a new generative prior as compared to the error BEAST2 would make when operating on a tree generated under the right tree prior (the twin tree).
From such results an user can estimate whether or not a new tree prior, tailored on the generative process, is needed. If this would indeed be the case, one could implement their own tree prior as a new module in BEAST2.
\giovanni{Maybe we can add a bit more on the nested sampling.}

@@

>>>>

\giovanni{describe winner}
\richel{agreed, after the figures also show it :-) }

@@

\giovanni{describe winner}
\richel{agreed, after the figures also show it :-) }

@@

\richel{check numbers}
\giovanni{
  I think that the current version of the discussion is too didascalic. 
  Most of what is said can be effectively summarized by table 5. 
  We can even add ESSes in it. 
}
\richel{
  Agreed, I disliked it as well
}
\giovanni{
  Another point is that we should understand what is 
  the message we want to deliver in this discussion. 
  Since we are only using one tree we are not in the position to claim anything, 
  because a single tree is not an effective sample. 
  I saw that in some occasions you refer to the RQ as examples. 
  This might be the right way to proceed, 
  as our purpose is just to illustrate how to run all the possible experiments 
  on a full tree distribution. 
  I think we should make it clearer throughout the entire manuscript. 
  I am also interested in hearing Rampal's opinion on this.
}

\giovanni{
  This is a first draft on how I think the discussion should be setup. 
  In this context there is no much need for the results in the table. 
  Let me know what you think about it.
}
\richel{
  I think that was the right way forward. 
  I do think the reader would like help interpreting the
  results of the example research questions as if these were
  proper experiments. I added that
}

In the four example research questions we showed how to use \verb;pirouette; 
to assess if BEAST2 is able, with a birth-death prior, 
to recover the original tree simulated by a Yule process. 
In principle any other, possibly much more complex, 
generative prior can be tested.
All tests were performed on a single original tree, 
hence not providing us enough statistical power to 
evaluate the level of accuracy of the inference.

However if the same procedure were repeated and performed on a distribution 
with a sufficient number of generative trees, 
it could actually constitute a quantitative assessment 
of the quality of the inference.
To give an idea of the insights \verb;pirouette; can give,
we pretend the results of the example research questions
to be obtained from thousands of replicates.
In that case, figure \ref{fig:example_1} shows 
the minimum error BEAST2 when the
generative model is known, which can serve as 
a minimum noise level assessment in a bigger experiment.  
Figure \ref{fig:example_2} shows that 
\richel{
  reword after new results with 6 taxa tree
} 
the best candidate model gives an error distribution similar to the unknown tree prior,
hinting that the candidate model would be just as good in inference.
Figure \ref{fig:example_3} shows that 
\richel{
  reword after new results with 6 taxa tree
} 
the twin model gives an error distribution similar to the unknown tree prior,
hinting that the shape of the phylogeny with unknown tree prior
does not effect the error made by BEAST2.
Figure \ref{fig:example_4} shows that 
\richel{
  reword after new results with 6 taxa tree
} 
the twin model and best candidate model give 
an error distribution similar to the unknown tree prior,
hinting that the shape of the phylogeny with unknown tree prior
does not effect the error made by BEAST2, 
and that the candidate model would be just as good in inference.

In conclusion, \verb;pirouette; can directly show 
the comparison between the results obtained 
on a new generative prior as compared to the error BEAST2 
would make when operating on a tree generated 
under a known, right and existing tree prior (the twin tree).
From such results an user can estimate whether or not a new tree prior, 
tailored on the generative process, is needed. 
If this would indeed be the case, 
one could implement their own tree prior as a new module in BEAST2.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h]
\centering
\begin{tabular}{ | r | l | l | l | l | }
\hline
\multirow{2}{*}{\textbf{Example}} & \multicolumn{2}{c|}{\textbf{True tree}} 
                                  & \multicolumn{2}{c|}{\textbf{Twin tree}} \\
\cline{2-5}
                                  & \textbf{median} & \textbf{range} & \textbf{median} & \textbf{range} \\
\hline
1  & $0.02$  & $0.0 - 0.27$  &         &               \\
2  & $0.02$  & $0.0 - 0.08$  &         &               \\
3  & $0.018$ & $0.0 - 0.1$   & $0.019$ & $0.0 - 0.11$  \\
4C & $0.22$  & $0.0 - 0.088$ & $0.009$ & $0.0 - 0.056$ \\
4G & $0.020$ & $0.0 - 0.075$ & $0.013$ & $0.0 - 0.052$ \\
\hline
\end{tabular}
\caption{
  \richel{redo}
  Results of the examples. 
  For the true and twin tree,
  the median and range (1st and 3rd quartile) of the errors 
  is shown. Experiment 4 is split in two rows
  by either the candidate ('C') or generative model ('G')
}
\label{tab:results}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
